---
title: "week 11"
format:
  html:
    embed-resources: true
editor: visual
---

## Homework

Estimate and interpret the coefficients of 7 models with a single two-group predictor. Use the following: 

-three binary models (one LPM, one log, one logit) 

-two “linear regressions” (Gaussian GLMs) 

-two count models

```{r}
#| message: false
#| warning: false
#| echo: false
#| cashe: false

library(gssr)
library(tidyverse)
library(tidyr)
```

## Variables

The predictor is one's education level, the outcome is one's first birth age.

```{r}
gss2018 <- gss_get_yr(2018)
fba <- gss2018 |>
  select(educ,agekdbrn) |>
  mutate(highschool = ifelse(educ >=12, 1, 0)) |>
  drop_na()
```

## Binary Models

First, we create binary outcomes.

```{r}
fbab <- fba |>
  mutate(latebirth = ifelse(agekdbrn >= 28, 1, 0)) 
```

### LPM

```{r}
lpm <- glm(latebirth ~ highschool,
           data = fbab,
           family = binomial(link = "identity"))
broom::tidy(lpm)
```

P(latebirth=1) = 0.106 + 0.198*highschool

People with high school degree increase the probability of having late birth by 19.8 percentage points. 95% confidence interval is (0.153, 0.243) 95% chances that people with high school degree increase the probability of having late birth by 15.3 to 24.3 percentage points in real population.

### Log

```{r}
logmodel <- glm(latebirth ~ highschool,
           data = fbab,
           family = binomial(link = "log"))
broom::tidy(logmodel)
```

log(P(latebirth=1)) = -2.242 + 1.053 *highschool

```{r}
exp(1.0532)
```

People with high school degree are 2.87 times more likely to have late birth age.

95% confidence interval is (`r round(1.0532-1.96*0.1897,3)`,`r round(1.0532+1.96*0.1897,3)`).

```{r}
exp(0.681)
exp(1.425)
```

95% chances are that People with high school degree are 1.98 to 4.16 times more likely to have late birth age.



### Logit

```{r}
logitmodel <- glm(latebirth ~ highschool,
           data = fbab,
           family = binomial(link = "logit"))
broom::tidy(logitmodel)
```

log($\frac{P(latebirth =1)}{(1-(P(latebirth = 1)}$) = -2.13 + 1.30\*highschool

```{r}
exp(1.30)
```

OR = 3.67. Having high school degree increases one's odds of having late birth age by 3.67 times.

95% confidence interval is (`r round(1.30-1.96*0.215,3)`,`r round(1.30+1.96*0.215,3)`).

```{r}
exp(0.879)
exp(1.721)
```

95 chances are that having high school degree increases one's odds of having late birth age by 2.41 to 5.59 times among true population.

## Linear Regression

Here we use first birth age rather than late birth.

```{r}
lmmodel <- lm(formula = agekdbrn ~ 1 + educ >=12, data = fba)
broom :: tidy(lmmodel)
```

First birth age = 20.8 + 3.86 \* highschool

### LMG

```{r}
Glm <- glm(agekdbrn ~ highschool,
           data = fba,
           family = gaussian(link = "identity"))
broom::tidy(Glm)
```

Firstbirthage = 21.1 + 3.78 * highschool

For Having high school degree is 3.78 years older for the first birth age compared to those not having high school degree.

95% confidence interval is (`r round(3.78-1.96*0.386,3)`,`r round(3.78+1.96*0.386,3)`) 95% chances that having high school degree is 3.02 to 4.54 years older for the fitst birth age compared to those not having high school degree.

### LogG

```{r}
Glog <- glm(agekdbrn ~ highschool,
           data = fba,
           family = gaussian(link = "log"))
broom::tidy(Glog)
```

log[P(agekdbrn)] = 3.05 + 0.165 * highschool

```{r}
exp(0.165)
exp(3.05)
```

The average first birth age for people without high school degree is 21.1. The expected age at first birth is about 18% higher for people with a high school education(21.1\*1.18=24.9) compared to those without.

95% confidence Interval is (`r round(0.165-1.96*0.0179,3)`,`r round (0.165+1.96*0.0179,3)`), 95% chances that the expected age at first birth is 13% to 22% higher for people with high school education compared to those without.

## Count Models
### LMP
```{r}
Plm <- glm(agekdbrn ~ highschool,
           data = fba,
           family = poisson(link = "identity"))
broom::tidy(Plm)
```

Firstbirthage = 21.1 + 3.78 \* highschool

For Having high school degree is 3.78 years older for the first birth age compared to those not having high school degree.

95% confidence interval is (`r round(3.78-1.96*0.322,3)`,`r round(3.78+1.96*0.322,3)`) 95% chances that having high school degree is 3.15 to 4.41 years older for the first birth age compared to those not having high school degree.

### LogP
```{r}
Plog <- glm(agekdbrn ~ highschool,
           data = fba,
           family = poisson(link = "log"))
broom::tidy(Plog)
```

```{r}
exp(0.165)
exp(3.05)
```

The average first birth age for people without high school degree is 21.1. The expected age at first birth is about 18% higher for people with a high school education(21.1*1.18=24.9) compared to those without.

95% confidence Interval is (`r round(0.165-1.96*0.0149,3)`,`r round (0.165+1.96*0.0149,3)`), 95% chances that the expected age at first birth is 14% to 19% higher for people with high school education compared to those without.



## Loglike

### Binary identity
lpm <- glm(latebirth ~ highschool,
           data = fbab,
           family = binomial(link = "identity"))

```{r}
d1 <- fbab |>
     mutate(predprob = predict(lpm, type = "response"),
            l = dbinom(latebirth, 1, predprob),
            ll = log(l))
sum(d1$ll)
logLik(lpm)
```
Manually calculate and function is -955.34

### Linear identity
Glm <- glm(agekdbrn ~ highschool,
           data = fba,
           family = gaussian(link = "identity"))

```{r}
d2 <- fba |> mutate(mean = predict(Glm),
                    sd = sd(Glm$residuals),
                    l2 = dnorm(agekdbrn,
                               mean = mean,
                               sd = sd),
                    ll2 = dnorm(agekdbrn,
                                mean = mean,
                                sd = sd,
                                log = TRUE))
sum(d2$ll2)
logLik(Glm)
```
Manually calculate and function is -5224.44                 

### Count Log
Plog <- glm(agekdbrn ~ highschool,
           data = fba,
           family = poisson(link = "log"))
```{r}
d3 <-  fba |> mutate(lambda = predict(Plog, type = "response"),
                    sd = sd(Plog$residuals),
                    l3 = dpois(agekdbrn,
                               lambda = lambda),
                    ll3 = dpois(agekdbrn,
                                lambda = lambda,
                                log = TRUE))
sum(d3$ll3)
logLik(Plog)
```
Manually calculate and function is -5189.72

Overall, Binary Identity has the largest result (closest to 0), thus it has better predictor of the correlation of having high school degree and late birth.